---
title: "Cleaning of Western Cape Dam Levels Data"
output: md_document
 
    
---

#Cleaning Western Cape Dam Level Data 

The following code is created for the purpose of extracting and cleaning data on the dam levels in the Western Cape (WC) province of South Africa. The data is provided by the WC provincial government.

The data is available from: https://web1.capetown.gov.za/web1/opendataportal/DatasetDetail?DatasetName=Dam+levels

##Load Dependencies

```{r, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
```

I use knitr::kable for the rendering of the markdown tables. This makes it easier to read in many formats, notably Github.

##Reading in the Data

```{r, message=FALSE, warning=FALSE}
DamLevels <- read_csv("Data/Raw_Data.csv")

DamLevels %>% 
  head(10) %>% 
  select(1:9) %>% 
  kable()
```

The data is not in a tidy form and so requires cleaning.

##Dates

A logical place to begin is with the date column as all the dam levels are aligned to it in wide format.

```{r}
DamLevelsDate<-DamLevels[5:nrow(DamLevels),1]    # The dates begin in the 5th row

DamLevelsDate %>% 
  head(10) %>% 
  kable()
```

The dates are parsed in as a character column at the outset. In order to parse the strings into a date format, I make use of the lubridate package.

```{r, message=FALSE, warning=FALSE}
library(lubridate)

DamLevelsDateClean <- DamLevelsDate %>% 
  rename(Date=`BULK WATER STORAGE`) %>%     # I rename the column to remove the full caps.
  mutate(Date=dmy(Date))

DamLevelsDateClean %>% 
  cbind(DamLevelsDate) %>% 
  head(10) %>% 
  kable()
```

The measurements are taken exactly one day apart. In the process of conducting the analysis however I found that there were some anomalies in the progression of the dates.

To check this, I perform some transformations on the data using dplyr. In particular I check to make sure that each date is exactly one day away from the date immediately above it.


```{r, message=FALSE, warning=FALSE}
DamLevelsDateClean %>% 
  mutate(initnum=1:nrow(DamLevelsDateClean)) %>% # I specify a new column of the initial row numbers for reference.
  mutate(DamLevelsDateLag=lag(DamLevelsDateClean$Date)) %>%   
  mutate(Diff=DamLevelsDateLag-Date) %>%  
  arrange(Diff)  %>% 
  head(10) %>% 
  kable()
```


This proves not to be the case. By some error, a few of the dates appear to be in the future. I managed to correct these by inspection.

```{r}
DamLevelsDateClean[1966,1]<-ymd("2017-05-19")
DamLevelsDateClean[1967,1]<-ymd("2017-05-20")
DamLevelsDateClean[1968,1]<-ymd("2017-05-21")
DamLevelsDateClean[1969,1]<-ymd("2017-05-22")
DamLevelsDateClean[1955,1]<-ymd("2017-05-08")

DamLevelsDateClean %>% 
  mutate(initnum=1:nrow(DamLevelsDate)) %>%   
  mutate(DamLevelsDateLag=lag(DamLevelsDateClean$Date)) %>%   
  mutate(Diff=DamLevelsDateLag-Date) %>%  
  arrange(Diff)  %>% 
  head(10) %>% 
  kable()
```


##Extracting Dam Names

The names of the various dams are given above the various columns and can be extracted in sequence. They are also changed to title case by means of the stringr package.



```{r, message=FALSE, warning=FALSE}
library(stringr)

seq<-seq(2,50,4)
Dams<-DamLevels[2,seq]    # The dam mnames are in the second row
Dams<-as.character(Dams)
Dams<-str_to_title(Dams)
Dams[4]<-"Voelvlei"   

Dams %>% 
  kable()
```

The name of Voelvlei dam usually contains a special character, which causes problems in various rendering envrioments. For simplicity it is replaced with the plain text version.


##Extracting the Storage Data

Each column represents a dam. The columns are evenly spaced and so can be extracted using a seq. The dam names (above) are used as the new column names.

```{r}
seq<-seq(3,51,4)

StorageWide<-DamLevels[5:nrow(DamLevels),seq]
colnames(StorageWide)<-Dams

StorageWide %>% 
  head(10) %>% 
  kable()
```

##Combine the Data

To combine the data, I bind the date column to the storage columns.

```{r}
StorageWide<-cbind(DamLevelsDateClean,StorageWide)

StorageWide %>% 
  head(10) %>% 
  kable()
```


##Convert the Data to Long Form

The data is still in wide form and so needs to be converted to long form. I do this using the reshape2 package.

The data from the storage column is still in character format and stil contains spaces. These are removed and the data is converted to double format.

```{r, message=FALSE, warning=FALSE}
library(reshape2)

StorageLong<-StorageWide %>% 
  melt(measure.vars = 2:ncol(StorageWide),variable.name = "Dam",value.name = "Storage") %>% 
  mutate(Storage=as.double(str_replace(Storage," ","")))

StorageLong %>% 
  head(10) %>% 
  kable()
```

##Extracting Capacity

Data on the capacities of the various dams line the top two rows of the dataset. They are evenly spaced and so can also be extracted by means of a sequence.

Again Voelvlei needs to be corrected. Again the characters need to be converted to doubles.

```{r}
seq<-seq(2,50,4)
Capacities<-DamLevels[1:2,seq]
Capacities<-data.frame(t(Capacities),stringsAsFactors = F)
Capacities[4,2]<-c("Voelvlei")


Capacities<-Capacities %>% 
  rename(Capacity=X1,Dam=X2) %>% 
  mutate(Capacity=as.double(str_replace(Capacity," ",""))) %>% 
  mutate(Dam=str_to_title(Dam))

Capacities  %>% 
  kable()
```

Calculate the total capacity:

```{r, message=FALSE, warning=FALSE}
TotalCapacity<-Capacities %>% 
  summarise(sum(Capacity)) %>% 
  as.numeric()

TotalCapacity %>% 
  kable()
```


#Merge Capacities with Long Data

To get the capacity data into the same dataframe as the storage data, I make use of the merge function. This automatically detects the appropriate column on which to merge.

```{r}
StorageLong<-StorageLong %>% 
  merge(Capacities)

StorageLong %>% 
  head(10) %>% 
  kable()
```

##Calculate Percentage Capacities

It is useful for plotting purposes to pre-calculate the percentage capacities.

Two are calculated. One is the percentage capcity of the dam in that particular row and the other is the  capcity of that dam as a percentage of the capacities of all the dams combined.

```{r}
StorageLong<-StorageLong %>% 
  mutate(PercentDamCapacity=Storage/Capacity) %>% 
   mutate(PercentTotalCapacity=Storage/TotalCapacity)

StorageLong %>% 
  head(10) %>% 
  kable()
```

```{r}
write.csv(file = "Data/Clean_WC_Dam_Levels.csv",StorageLong,row.names = F)
```

