---
title: "Cleaning Western Cape Dam Levels Data"
output: md_document
    
---


The following code is created for the purpose of extracting and cleaning data on the dam levels in the Western Cape (WC) province of South Africa. The data is provided by the WC provincial government.

The data is available from: https://web1.capetown.gov.za/web1/opendataportal/DatasetDetail?DatasetName=Dam+levels

##Load Dependencies

```{r}
library(readr)
library(magrittr)
```

##Reading in the Data

```{r}
DamLevels <- read_csv("~/Desktop/Dam levels update 2012-2017.csv")
DamLevels %>% 
  head(20)
```

The data is not in a tidy form and so requires cleaning.

#Dates

It is logical to begin with the DATE column and join it to the remaining columns afterwards.

```{r}
DamLevelsDate<-DamLevels[5:nrow(DamLevels),1]
DamLevelsDate %>% 
  head(20)
```

As can be seen the dates are parsed in as a character column at the outset. In order to parse the strings into a date format, I make use of the lubridate package.

```{r}
library(lubridate)

colnames(DamLevelsDate)<-c("Date")    # I rename the column to remove the all caps.
DamLevelsDate$Date<-dmy(DamLevelsDate$Date)

DamLevelsDate %>% 
  head(20)
```

As can be seen, measurements are taken exactly one day apart. To check this, I perform some transformations on the data using dplyr. In particular I check to make sure that each date is exactly one day different from the date immediately above it.

```{r}
library(dplyr)

DamLevelsDate %>% 
  mutate(initnum=1:nrow(DamLevelsDate)) %>%   
  mutate(DamLevelsDateLag=lag(DamLevelsDate$Date)) %>%   
  mutate(Diff=DamLevelsDateLag-Date) %>%  
  arrange(Diff)  %>% 
  head(20)
```

This proves not to be the case. By some error, a few of the dates appear to be in the future. I managed to correct these by inspection.

```{r}
DamLevelsDate[1966,1]<-ymd("2017-05-19")
DamLevelsDate[1967,1]<-ymd("2017-05-20")
DamLevelsDate[1968,1]<-ymd("2017-05-21")
DamLevelsDate[1969,1]<-ymd("2017-05-22")
DamLevelsDate[1955,1]<-ymd("2017-05-08")

DamLevelsDate %>% 
  mutate(initnum=1:nrow(DamLevelsDate)) %>%   
  mutate(DamLevelsDateLag=lag(DamLevelsDate$Date)) %>%   
  mutate(Diff=DamLevelsDateLag-Date) %>%  
  arrange(Diff)  %>% 
  head(20)
```

##Extracting Dam Names

The names of the various Dams are given above the various columns and can be extracted in sequence. They are also changed to title case by means of the stringr package.

```{r}
library(stringr)
seq<-seq(2,50,4)
Dams<-DamLevels[2,seq]
Dams<-as.character(Dams)
Dams<-str_to_title(Dams)
Dams
```

##Extracting the Storage Data

Each column represents a dam. The columns are evenly spaced and so can be extracted using a seq. The Dam names (above) are used as the new column names.

```{r}
seq<-seq(3,51,4)
StorageWide<-DamLevels[5:nrow(DamLevels),seq]
colnames(StorageWide)<-Dams
StorageWide %>% 
  head(20)
```

##Combine the Data

To combine the data, I bind the Date column to the storage columns.

```{r}
StorageWide<-cbind(DamLevelsDate,StorageWide)
StorageWide %>% 
  head(20)
```

##Convert the Data to Long Form

```{r}
library(reshape2)
StorageLong<-StorageWide %>% 
  melt(measure.vars = 2:ncol(StorageWide),variable.name = "Dam",value.name = "Storage") %>% 
  mutate(Storage=as.double(str_replace(Storage," ","")))

StorageLong %>% 
  head(20)
```
##Extracting Capacity

```{r}
seq<-seq(2,50,4)
Capacities<-DamLevels[1:2,seq]
Capacities<-data.frame(t(Capacities))

Capacities<-Capacities %>% 
  rename(Capacity=X1,Dam=X2) %>% 
  mutate(Capacity=as.double(str_replace(Capacity," ",""))) %>% 
  mutate(Dam=str_to_title(Dam))
Capacities  
```

Calculate total

```{r}
TotalCapacity<-Capacities %>% 
  summarise(sum(Capacity)) %>% 
  as.numeric()

TotalCapacity
```


#Merge Capacities with Long Data

```{r}
StorageLong<-StorageLong %>% 
  merge(Capacities)
StorageLong %>% 
  head(20)
```

#Calculate Percentage Capacities

```{r}
StorageLong<-StorageLong %>% 
  mutate(PercentDamCapacity=Storage/Capacity) %>% 
   mutate(PercentTotalCapacity=Storage/TotalCapacity)
StorageLong %>% 
  head(20)
```

```{r}
write.csv(file = "/Users/jameslairdsmith/Google Drive/Applications/Github/WC_Dam_Levels/Clean_WC_Dam_Levels.csv",StorageLong)
```

